#Getting and Cleaning Data Quiz 1

###Question 1
The American Community Survey distributes downloadable data about United States communities. Download the 2006 microdata survey about housing for the state of Idaho using download.file() from here: 
```{html}
https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2Fss06hid.csv 
```
and load the data into R. The code book, describing the variable names is here: 
```{html}
https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2FPUMSDataDict06.pdf 
```
How many properties are worth $1,000,000 or more?
```{r}
download.file("https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2Fss06hid.csv",destfile="quiz1.csv")
q1data<-read.csv("quiz1.csv",header=TRUE)
length(which((q1data[,"VAL"]==24)))
```


###Question 2
Use the data you loaded from Question 1. Consider the variable FES in the code book. Which of the "tidy data" principles does this variable violate?
```{r}
head(q1data$FES)
str(q1data$FES)
```
Checking the code book, it combines two variables.


###Question 3
Download the Excel spreadsheet on Natural Gas Aquisition Program here: 
```{html}
https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2FDATA.gov_NGAP.xlsx 
```
Read rows 18-23 and columns 7-15 into R and assign the result to a variable called:
```{r}
dat 
````
What is the value of:
```{r}
sum(dat$Zip*dat$Ext,na.rm=T)
```
original data source:
```{html} http://catalog.data.gov/dataset/natural-gas-acquisition-program)
```
```{r}
library(rjava)
library(xlsx)
download.file("https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2FDATA.gov_NGAP.xlsx",destfile="NatGas.csv")
natGas<-read.csv("NatGas.csv")
sum(dat$Zip*dat$Ext,na.rm=T)
```


###Question 4
Read the XML data on Baltimore restaurants from here: 
```{html} https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2Frestaurants.xml 
```
How many restaurants have zipcode 21231?
```{r}
library(XML)
fileUrl<-"https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2Frestaurants.xml"
download.file(fileUrl,destfile="quiz1.xml")
xmlFile<-xmlParse("quiz1.xml")
xmlTop<-xmlRoot(xmlFile)
xmlName(xmlTop)
xmlSize(xmlTop)
xml<-xmlToDataFrame(xmlFile)
xmlToList(xmlFile)

xmlTreeParse(xmlFile,useInternal=TRUE)
root<-xmlRoot(xmlFile)
xmlName(rootNode)
```


###Question 5
The American Community Survey distributes downloadable data about United States communities. Download the 2006 microdata survey about housing for the state of Idaho using download.file() from here: 
```{html}
https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2Fss06pid.csv 
```
using the fread() command load the data into an R object
```{r}
DT 
```
Which of the following is the fastest way to calculate the average value of the variable
pwgtp15 
broken down by sex using the data.table package?
```{r}
sapply(split(DT$pwgtp15,DT$SEX),mean)
```
